### 2018-08-13 doc/manuscript.Rmd
Updates

---
title: Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 12pt
mainfont: 'Helvetica'
---

@@

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven by eukaryotic viruses or by bacteriophages. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggests a more complex relationship between the bacterial and phage communities.

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Based on our findings, we have developed a conceptual model to be tested in future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer. The dsDNA virome of colorectal cancer was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating. We hypothesize that the process begins with broadly infectious phages in the colon lysing, and thereby disrupting, the existing bacterial communities. This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria establish themselves in the epithelium, secondary opportunistic bacteria are able to adhere to the founders, colonize, and establish a biofilm. Phages may play a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then be able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria enter their oncogenic synergy with the epithelium, we conjecture that the phages continue mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities this model will guide future research investigations of the role the virome plays colorectal cancer.

Our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system can be operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer. Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group or no-DNA negative control. B) Sequence yield following quality control including quality score filtering and human decontamination. Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis.*\label{qualcontrol}]()

>>>>

---
title: The Diagnostic Potential & Interactive Dynamics of the Colorectal Cancer Virome
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 12pt
mainfont: 'Helvetica'
---

@@

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven by eukaryotic (human) viruses or by bacteriophages. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the bacteriophage signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria.

Given these findings, we posited that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages that were important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Based on our findings, we have developed a conceptual model to be tested in our future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer. The dsDNA virome of colorectal cancer was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into a current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating. We hypothesize that the process begins with broadly infectious phages in the colon lysing, and thereby disrupting, the existing bacterial communities. This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria establish themselves in the epithelium, secondary opportunistic bacteria are able to adhere to the founders, colonize, and establish a biofilm. Phages may play a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then be able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria enter their oncogenic synergy with the epithelium, we conjecture that the phages continue mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities this model will guide future research investigations of the role the virome plays colorectal cancer.

Our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system may operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. To better understand this system, future studies will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling. Overall, this study provides a conceptual foundation to direct future characterization of the colorectal cancer microbiome at the functional, mechanistic level.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group or no-DNA negative control. B) Sequence yield following quality control including quality score filtering and human decontamination. Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis. After rarefaction and removal of samples with less than $10^6$ reads, 27 healthy, 28 cancerous, 27 adenomatous, and 3 negative control samples remained.*\label{qualcontrol}]()

### 2018-08-13 revfiles/internalComments.rmd
Updates

<CHANGE LEGEND AND RERENDER>

Sorry I don't see the discordance between the figures. Maybe we are looking at different versions?

Love the rewording, that is much better!

@@

<ADD DASHED LINE TO LEGEND>

@@

## COMMENT 4

We need to be careful about getting too caught up in the specific "important" phages because none of them are really that important. The importance levels are small, and multiple times less than what was observed by bacteria. So it's probably not a single phage involved. Also these are representative sequences from OVUs, and are not really meant to provide specific resolution, as described for the PLOS comp bio paper.

@@

Thanks, the intent was not to have it read as the only alternative hypothesis. We can add more, and point out that those are also not the only alternatives, because I think we could continue hypothesizing about this and coming up with more options.

<FIX THE TEXT>

@@

I like this, although I would be hesitant to say non-linear patterns because, while it excells at complex non-linear relationships, it should be able to detect them as well.

@@

Thanks for the feedback. I'll also try to make it sound less dodgy. The intent was to do the re-analyses and textual revisions that were warranted.

@@

>>>>

Sorry I don't see the discordance between the figures. Maybe we are looking at different versions? Regardless it should be fixed now.

Rewording was also great.

@@

@@

Done.

## COMMENT 4-A

This is great feedback, thanks. I went through and made some revisions to hopefully make this clearer.

## COMMENT 4-B

I think we need to be careful about getting too caught up in the specific "important" phages because none of them are really all that important conpmared to the others. The importance levels are close, and multiple times less than what was observed by bacteria. So it's probably not a single phage involved and I don't think it really makes sense to get that caught up in a single phage (a point I probably didn't hit hard enough in the text). Also it is still important to note that these are representative sequences from OVUs, and are not really meant to provide specific resolution, as described for the PLOS comp bio paper. It could be a good idea to really dig into this if we had more compelling data (i.e. similar importance signal to the fusobacterium case on the CRC bacterial side of things) but here we are really only looking at community wide signatures without diving into any given specific elements.

@@

Thanks, the intent was not to have it read as the only alternative hypothesis. We can add more, and point out that those are also not the only alternatives, because I think we could continue hypothesizing about this and coming up with more options. I'll add in these suggestions and make it clearer that there are many options to explore.

@@

I like this, although I am hesitant to focus on non-linear patterns in the language because, while it excells at complex non-linear relationships, it should be able to a good job detecting linear relationships as well.

@@

Thanks for the feedback. I'll also try to make it sound less dodgy.

@@

This is great and incorporated the changes.

I agree that it should be self evident but this is something that some of my colleagues, who are not familiar with phages, appreciated hearing.

These are also great points about how phages will interact with their bacterial hosts, and their lysogenic replication cycles. These will be important as we work to understand the system.
